R for Clinical Study Reports and Submission

### 2021-11-15 tlf-baseline.Rmd

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize critical demographic and baseline characteristics of the patients
in Section 11.2, Demographic and Other Baseline Characteristics.

In this chapter, we illustrate how to create a simplified
baseline characteristics table in a study.

@@

We start to define the format of the output. We highlight items that
are not discussed in previous discussion.

`text_indent_first` and `test_indent_left` are used to control the indent
space of text. They are helpful when you need to control the white space
of a long sentence. For example, "AMERICAN INDIAN OR ALASKA NATIVE"
in the table.

@@

tbl_base[-1, ] %>%
  rtf_title(
    "Participant Baseline Characteristics",
    "(All Participants Randomized)"
  ) %>%
  rtf_colheader(colheader1,
    col_rel_width = rel_width
  ) %>%
  rtf_colheader(colheader2,
    border_top = "",
    col_rel_width = rel_width
  ) %>%
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 4)),
    text_indent_first = -240,
    text_indent_left = 180
  ) %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_base.rtf")
```

@@

In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows.

- Step 1: Read the data set.
- Step 2: Use `table1::table1()` to get the baseline characteristics table.
- Step 3: Transfer the output in Step 2 into a data frame that only contains ASCII characters.
- Step 4: Define the format of the RTF table by using the R package `r2rtf`.

>>>>

Following [ICH E3 guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize critical demographic and baseline characteristics of the participants
in Section 11.2, Demographic and Other Baseline Characteristics.

In this chapter, we illustrate how to create a simplified
baseline characteristics table for a study.

@@

We define the format of the output. We highlight items that
are not discussed in previous discussion.

`text_indent_first` and `test_indent_left` are used to control the indent
space of text. They are helpful when you need to control the white space
of a long phrase "AMERICAN INDIAN OR ALASKA NATIVE"
in the table provide an example.

@@

tbl_base[-1, ] %>%
  rtf_title(
    "Baseline Characteristics of Participants",
    "(All Participants Randomized)"
  ) %>%
  rtf_colheader(colheader1,
    col_rel_width = rel_width
  ) %>%
  rtf_colheader(colheader2,
    border_top = "",
    col_rel_width = rel_width
  ) %>%
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 4)),
    text_indent_first = -240,
    text_indent_left = 180
  ) %>%
  rtf_encode() %>%
  write_rtf("tlf/tlf_base.rtf")
```

@@

In conclusion, the procedure to generate demographic and baseline characteristics table is summarized as follows:

- Step 1: Read the data set.
- Step 2: Use `table1::table1()` to get the baseline characteristics table.
- Step 3: Transfer the output from Step 2 into a data frame that only contains ASCII characters.
- Step 4: Define the format of the RTF table by using the R package `r2rtf`.

### 2021-11-15 tlf-efficacy.Rmd

For efficacy analysis, only the change from baseline glucose data at week 24 is analyzed.

@@

First, both the population and the data in scope are selected.
The analysis is done on the efficacy population, identified by `EFFFL == "Y"`, and
all records post baseline (`AVISITN >= 1`) and on or before Week 24 (`AVISITN <= 24`).
Here the variable `AVISITN` is the numerical analysis visit.
For example, if the analysis visit is recorded as "Baseline", i.e., `AVISIT = Baseline`, then `AVISITN = 0`;
if the analysis visit is recorded as "Week 24", i.e., `AVISIT = Week 24`, then `AVISITN = 24`;
if the analysis visit is blank, then `AVISITN` is also blank.
We will discuss these missing values in Section 6.4.

@@

To prepare the report data frame, we created a few helper functions
by using the `fmt_num()` function defined in Chapter \@ref(population).

- Format Estimators

@@

- Format Confidence Interval

@@

- Format P-Value

@@

First the observed data at Baseline and Week 24 is summarized:

@@

Also the observed change from baseline glucose at Week 24 is summarized:

@@

In clinical trials, missing data is inevitable.
In this study, there are also missing values in glucose data.

@@

For simplicity and illustration purpose,
we use the last observation carried forward (LOCF) approach to handle missing data.
LOCF approach is a single imputation approach that is **not recommended**
in real studies.
Interested readers can find more discussion on missing data approaches in the book:
[The Prevention and Treatment of Missing Data in Clinical Trials](https://www.ncbi.nlm.nih.gov/books/NBK209904/pdf/Bookshelf_NBK209904.pdf).

@@

The `emmeans` R package is used to obtain
within and between group least square mean (LS mean)

@@

`t11`, `t12` and `t13` are combined to get the first part of the report data

@@

Special characters `^` and `_` are used to define superscript and subscript of text. And `{}` is to define the part that will be impacted.
For example, `{^a}` provide a superscript `a` for footnote notation.
`r2rtf` also supports most LaTeX characters.
Examples can be found on the
[`r2rtf` get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
The `text_convert` argument in `r2rtf_*()` functions controls whether convert special characters.

@@

we also use `r2rtf` to prepare the table format for `t2`

@@

Finally we combine the two parts to get the final table using `r2rtf`.
This is achieved by providing a list of `t1_rtf` and `t2_rtf` as input for
`rtf_encode`.

@@

- Step 1: Read the data into R, i.e., `adsl` and `adlb`.
- Step 2: Define the analysis dataset. In this example, we define two analysis datasets. The first dataset is the efficacy population (`gluc`). The second dataset is the collection of all records post baseline and on or before week 24 (`ana`).
- Step 3: Impute the missing values. In our example, we name the `ana` dataset after imputation as `ana_locf`.
- Step 4: Summarize `gluc`, i.e., calculate the mean and standard derivation, and then format it into an RTF table.
- Step 5: Summarize `ana_locf`, i.e., calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
- Step 6: Combine the outputs in step 4 and step 5 by rows.

>>>>

In this chapter, we illustrate how to generate an efficacy table for a study.
For efficacy analysis, only the change from baseline glucose data at week 24 is analyzed.

@@

First, both the population and the data in scope are selected.
The analysis is done on the efficacy population, identified by `EFFFL == "Y"`, and
all records post baseline (`AVISITN >= 1`) and on or before Week 24 (`AVISITN <= 24`).
Here the variable `AVISITN` is the numerical analysis visit.
For example, if the analysis visit is recorded as "Baseline" (i.e., `AVISIT = Baseline`), 
`AVISITN = 0`;
if the analysis visit is recorded as "Week 24" (i.e., `AVISIT = Week 24`), `AVISITN = 24`;
if the analysis visit is blank, `AVISITN` is also blank.
We will discuss these missing values in Section 6.4.

@@

To prepare the report, we create a few helper functions
by using the `fmt_num()` function defined in Chapter \@ref(population).

- Format estimators

@@

- Format confidence interval

@@

- Format p-value

@@

First the observed data at Baseline and Week 24 are summarized using code below:

@@

Also the observed change from baseline glucose at Week 24 is summarized using code below:

@@

In clinical trials, missing data is inevitable.
In this study, there are missing values in glucose data.

@@

For simplicity and illustration purpose,
we use the last observation carried forward (LOCF) approach to handle missing data.
LOCF approach is a single imputation approach that is **not recommended**
in real application.
Interested readers can find more discussion on missing data approaches in the book:
[The Prevention and Treatment of Missing Data in Clinical Trials](https://www.ncbi.nlm.nih.gov/books/NBK209904/pdf/Bookshelf_NBK209904.pdf).

@@

The `emmeans` R package is used to obtain
within and between group least square (LS) mean

@@

`t11`, `t12` and `t13` are combined to get the first part of the report table

@@

Special characters `^` and `_` are used to define superscript and subscript of text. And `{}` is to define the part that will be impacted.
For example, `{^a}` provides a superscript `a` for footnote notation.
`r2rtf` also supports most LaTeX characters.
Examples can be found on the
[`r2rtf` get started page](https://merck.github.io/r2rtf/articles/r2rtf.html#special-character).
The `text_convert` argument in `r2rtf_*()` functions controls whether to convert special characters.

@@

We also use `r2rtf` to prepare the table format for `t2`

@@

Finally, we combine the two parts to get the final table using `r2rtf`.
This is achieved by providing a list of `t1_rtf` and `t2_rtf` as input for
`rtf_encode`.

@@

- Step 1: Read the data (i.e., `adsl` and `adlb`) into R.
- Step 2: Define the analysis dataset. In this example, we define two analysis datasets. The first dataset is the efficacy population (`gluc`). The second dataset is the collection of all records post baseline and on or before week 24 (`ana`).
- Step 3: Impute the missing values. In this example, we name the `ana` dataset after imputation as `ana_locf`.
- Step 4: Calculate the mean and standard derivation of efficacy endpoint (i.e., `gluc`), and then format it into an RTF table.
- Step 5: Calculate the pairwise comparison by ANCOVA model, and then format it into an RTF table.
- Step 6: Combine the outputs from steps 4 and 5 by rows.

### 2021-11-15 tlf-population.Rmd
rename `adam_data/` as `data-adam/`

Following [ICH E3
guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize which participants were included in each efficacy
analysis in Section 11.1, Data Sets Analysed.

@@

In this chapter, we illustrate how to create a summary table for
the analysis population in a study.

@@

-   USUBJID: Unique subject identifier
-   ITTFL: Intent-to-treat population flag
-   EFFFL: Efficacy population flag
-   SAFFL: Safty population flag

@@

Before we write the analysis code, let's discuss the possibility of reusing R
code by writing some toy helper functions.

@@

-   Format the percentages using the `formatC()` function.
-   Calculate the numbers and percentages by groups.

We create two ad-hoc functions and apply them to all the rest of the
tables.

To format numbers and percentages, we created a function called
`fmt_num()`. It is a very simple function wrapping `formatC()`.

@@

For example, we can compare the two versions of code to format the
percentage used in \@ref(disposition) and `fmt_num`.

@@

With the helper functions `count_by`, we can easily prepare a report
dataset as

@@

Now we combine individual rows into the whole table for reporting
purposes. `tbl_pop` is used as input for `r2rtf` to create the final
report.

@@

We start to define the format of the output.

```{r}
rel_width <- c(2, rep(1, 3))
colheader <- " | Placebo | Xanomeline line Low Dose| Xanomeline line High Dose"
tbl_pop %>%
  # Table title
  rtf_title(
    "Participants Accounting in Analysis Population",
    "(All Participants Randomized)"
  ) %>%
  # First row of column header
  rtf_colheader(colheader,
    col_rel_width = rel_width
  ) %>%
  # Second row of column header
  rtf_colheader(" | n (%) | n (%) | n (%)",
    border_top = "",
    col_rel_width = rel_width
  ) %>%
  # Table body
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 3))
  ) %>%
  # Encoding RTF syntax
  rtf_encode() %>%
  # Save to a file
  write_rtf("tlf/tbl_pop.rtf")
```

@@

The procedure to generate an analysis population table can be summarized
as follow:

- Step 1: Read data into R, i.e., `adsl`.
- Step 2: Bind the counts/percentages of the ITT population, the
  efficacy population, and the safety population by row using the
  `count_by()` function.
- Step 3: Format the output in Step 2 by `r2rtf`.

>>>>

Following [ICH E3
guidance](https://database.ich.org/sites/default/files/E3_Guideline.pdf),
we need to summarize the number of participants included in each efficacy
analysis in Section 11.1, Data Sets Analysed.

@@

In this chapter, we illustrate how to create a summary table for
the analysis population for a study.

@@

-   USUBJID: unique subject identifier
-   ITTFL: intent-to-treat population flag
-   EFFFL: efficacy population flag
-   SAFFL: safty population flag

@@

Before we write the analysis code, let's discuss the possibility of reusing R
code by writing helper functions.

@@

-   Format the percentages using the `formatC()` function.
-   Calculate the numbers and percentages by treatment arm.

We create two ad-hoc functions and use them to create the
tables in the rest of this book.

To format numbers and percentages, we create a function called
`fmt_num()`. It is a very simple function wrapping `formatC()`.

@@

For example, we can compare the two versions of code to format the
percentage used in Chapter \@ref(disposition) and `fmt_num`.

@@

With the helper function `count_by`, we can easily prepare a report
dataset as

@@

Now we combine individual rows into one table for reporting
purpose. `tbl_pop` is used as input for `r2rtf` to create the final
report.

@@

We define the format of the output using code below.

```{r}
rel_width <- c(2, rep(1, 3))
colheader <- " | Placebo | Xanomeline line Low Dose| Xanomeline line High Dose"
tbl_pop %>%
  # Table title
  rtf_title(
    "Summary of Analysis Sets", ## "Participants Accounting in Analysis Population" is a weird title for me. Merck uses this table title for CSR? 
    "(All Participants Randomized)"
  ) %>%
  # First row of column header
  rtf_colheader(colheader,
    col_rel_width = rel_width
  ) %>%
  # Second row of column header
  rtf_colheader(" | n (%) | n (%) | n (%)",
    border_top = "",
    col_rel_width = rel_width
  ) %>%
  # Table body
  rtf_body(
    col_rel_width = rel_width,
    text_justification = c("l", rep("c", 3))
  ) %>%
  # Encoding RTF syntax
  rtf_encode() %>%
  # Save to a file
  write_rtf("tlf/tbl_pop.rtf")
```

@@

The procedure to generate an analysis population table can be summarized
as follows:

- Step 1: Read data (i.e., `adsl`) into R.
- Step 2: Bind the counts/percentages of the ITT population, the
  efficacy population, and the safety population by row using the
  `count_by()` function.
- Step 3: Format the output from Step 2 using `r2rtf`.
